Highly Available Transactions: Virtues and Limitations

### 2013-04-01 writeup/model.tex
killed the hierarchy stuff

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage
  systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we consider partially
  replicated systems.} a server that can handle an operation for a
given data item is called a \textit{replica} for that item. While this
standard model is useful, there are several other cases we should
consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of operations will persist in the event of
server failures, clients may wish to wait until operations reach more
than one server before they consider the operations
completed. However, according to our definition of high availability,
server-fault tolerance is unavailable. To allow flexibility in the
number of participating servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing \textit{high availability} ($1$-availability), a system with
$2$-availability will be \textit{less available}, or require more
messaging and possibly higher latency and may experience more
unsuccessful operations in the presence of network and server
failures.  As is commmon in standard consensus algorithms like Paxos,
clients may want to contact a majority of $N$ servers in a
system. This \textit{majority availability} (a special case of
$K$-availability with $K=\lceil \frac{N}{2} \rceil$) is \textit{more
  available} than full $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness entails fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a server that has observed all of its prior operations, it eventually
receives a response, even in the presence of indefinitely long
partitions. As above, it may be useful for a client to achieve
stickiness \textit{and} contact many replicas (i.e., for fault
tolerance), so we generalize sticky availability to \textit{$K$-sticky
  availability}, which provides sticky availability with $K$ distinct
servers. A client may choose to become $1$-sticky available by acting
as a server itself; for example, a client might cache its reads and
writes~\cite{bolton, sessionguarantees, swift}. However, a single client
cannot achieve $K$-sticky availability with $K>1$ without contacting
other replicas. Moreover, ``sticking'' all clients to a designated set
of servers (i.e., master availability) is a special case of sticky
availability and is ``less available'' than the equivalent
non-mastered sticky availability.

\begin{figure}
\centering
\begin{tikzpicture}[scale=0.8]
  \tikzstyle{every node}=[font=\small]
 \node[draw=none,fill=none] (1) at (0,0) {$1$-availability (high availability)}; 
 \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
 \node[draw=none,fill=none] (m-1) at (0, 2) {\ldots}; 
 \node[draw=none,fill=none] (m) at (0, 3) {majority availability}; 
 \node[draw=none,fill=none] (m+1) at (0, 4) {\ldots}; 
 \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
 \node[draw=none,fill=none] (1s) at (4, 1.5) {$1$-sticky availability};
 \node[draw=none,fill=none] (2s) at (4, 2.5) {$2$-sticky availability};
 \node[draw=none,fill=none] (n2s) at (4, 3.5) {\ldots};
 \node[draw=none,fill=none] (n1s) at (4, 4.5) {$(N$$-$$1)$-sticky availability};

 \draw [->] (1) -- (2);
 \draw [->] (2) -- (m-1);
 \draw [->] (m-1) -- (m);
 \draw [->] (m) -- (m+1);
 \draw [->] (m+1) -- (n);
 \draw [->] (1) -- (1s);
 \draw [->] (1s) -- (m);
 \draw [->] (2s) -- (m+1);
 \draw [->] (1s) -- (2s);
 \draw [->] (2) -- (2s);
 \draw [->] (2s) -- (n2s);
 \draw [->] (n2s) -- (n1s);
 \draw [->] (n1s) -- (n);

\end{tikzpicture}
\caption{Hierarchy of replica availability levels for $N>3$ servers.}
\label{fig:availability-order}
\end{figure}

We show a hierarchy of replica availability levels in
Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
$K$-availability, but $K$-sticky availability is incomparable with
$(K+1)$-availability. $N$-sticky availability is equivalent to
$N$-availability, while majority availability subsumes $1$-sticky
availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
availability subsumes $K$-sticky availability. We have omitted
discussion of operation-specific availability levels (e.g.,
$N$-availability for writes and $1$-availability for reads in a
write-all, read-one data store), system membership changes, or
heterogeneous replicas (e.g., servers in a local and remote
datacenters) but believe there are several avenues for further
taxonomization.

@@

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally
aborts~\cite{hat-hotos}. For example, a system will violate
transactional $2$-availability if a client that can contact two
servers for each of the data items in its transaction does not
eventually commit in the absence of internal aborts.

>>>>

Traditionally, a system provides {\em  availability} if every user that
can contact a correct server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage
  systems~\cite{dynamo, pnuts, bigtable, spanner, hstore}, we consider partially
  replicated systems.} a server that can handle an operation for a
given data item is called a \textit{replica} for that item. While this
standard model is useful, there are several other cases we should
consider.

Some algorithms assume a model in which the client always contacts the
same server across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness entails fate-sharing between the client and its 
server: if a client's sticky server fail, or, equivalently, if the
client is partitioned from its sticky server, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a correct server that has observed all of its prior operations, it eventually
receives a response, even in the presence of indefinitely long
partitions. A client may choose to become sticky available by acting
as a server itself; for example, a client might cache its reads and
writes~\cite{bolton, sessionguarantees, swift}.

Availability implies sticky availability, since availability does not
require that the contacted server has seen all the previous operations
of a client, whereas sticky availability is the special case in which
the system only guarantees correctness if the contacted server has
seen all the prior operations from the client.

%% \begin{figure}
%% \centering
%% \begin{tikzpicture}[scale=0.8]
%%   \tikzstyle{every node}=[font=\small]
%%  \node[draw=none,fill=none] (1) at (0,0) {$1$-availability (high availability)}; 
%%  \node[draw=none,fill=none] (2) at (0,1) {$2$-availability}; 
%%  \node[draw=none,fill=none] (m-1) at (0, 2) {\ldots}; 
%%  \node[draw=none,fill=none] (m) at (0, 3) {majority availability}; 
%%  \node[draw=none,fill=none] (m+1) at (0, 4) {\ldots}; 
%%  \node[draw=none,fill=none] (n) at (0, 5) {$N$-availability}; 
%%  \node[draw=none,fill=none] (1s) at (4, 1.5) {$1$-sticky availability};
%%  \node[draw=none,fill=none] (2s) at (4, 2.5) {$2$-sticky availability};
%%  \node[draw=none,fill=none] (n2s) at (4, 3.5) {\ldots};
%%  \node[draw=none,fill=none] (n1s) at (4, 4.5) {$(N$$-$$1)$-sticky availability};

%%  \draw [->] (1) -- (2);
%%  \draw [->] (2) -- (m-1);
%%  \draw [->] (m-1) -- (m);
%%  \draw [->] (m) -- (m+1);
%%  \draw [->] (m+1) -- (n);
%%  \draw [->] (1) -- (1s);
%%  \draw [->] (1s) -- (m);
%%  \draw [->] (2s) -- (m+1);
%%  \draw [->] (1s) -- (2s);
%%  \draw [->] (2) -- (2s);
%%  \draw [->] (2s) -- (n2s);
%%  \draw [->] (n2s) -- (n1s);
%%  \draw [->] (n1s) -- (n);

%% \end{tikzpicture}
%% \caption{Hierarchy of replica availability levels for $N>3$ servers.}
%% \label{fig:availability-order}
%% \end{figure}

%% We show a hierarchy of replica availability levels in
%% Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
%% $K$-availability, but $K$-sticky availability is incomparable with
%% $(K+1)$-availability. $N$-sticky availability is equivalent to
%% $N$-availability, while majority availability subsumes $1$-sticky
%% availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
%% availability subsumes $K$-sticky availability. We have omitted
%% discussion of operation-specific availability levels (e.g.,
%% $N$-availability for writes and $1$-availability for reads in a
%% write-all, read-one data store), system membership changes, or
%% heterogeneous replicas (e.g., servers in a local and remote
%% datacenters) but believe there are several avenues for further
%% taxonomization.

@@

We say that a system provides \textit{transactional availability} if,
given replica availability for every data item in a transaction, the
transaction eventually commits or internally
aborts~\cite{hat-hotos}. For example, a system will violate
transactional availability if a client that can contact two
servers for each of the data items in its transaction does not
eventually commit in the absence of internal aborts.
